---
title: "Using tidymodels package - Part 2 - New York city flights"
format:
  html:
    code-fold: show
    code-tools: true
---

## Code setup
```{r}
#| output: false
library(here)
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidymodels)
library(nycflights13)
```

```{r}
#| output: FALSE
theme_set(theme_gray(base_size = 12)) + 
    theme_update(
        legend.position = "bottom",
        strip.background = element_blank(),
    )
```

## Load and explore the data
```{r}
dat <- flights |>
    filter(!is.na(arr_delay)) |>
    mutate(flight_date = as_date(time_hour),
           arr_delay_fct = factor(if_else(arr_delay >= 30, "delayed", "ontime"))) |>
    #inner_join(weather, by=c("origin", "time_hour")) |>
    select(dep_time, flight, origin, dest, air_time, distance, 
           carrier, flight_date, arr_delay, arr_delay_fct, time_hour) |>
    mutate_if(is.character, as.factor)
```

**What proportion of flights is delayed?**

```{r}
dat |>
  count(arr_delay_fct) |>
  mutate(proportion_of_flights = n / sum(n))
```

**Understanding interactions between ariline, flight destinations, and flight distance**

```{r}
sort(table(dat$dest), decreasing = TRUE)
```

```{r}
sort(table(dat$carrier), decreasing = TRUE)
```

The graph below shows that, perhaps unsuprisingly, there is a relationship between airline and flight distance...

```{r}
#| fig.height: 10
#| fig.width: 16

dat |>
 ggplot() +
 geom_histogram(aes(x=distance, fill=arr_delay_fct), bins=70) +
 facet_wrap(~carrier) +
 scale_fill_viridis_d(option = "plasma", end=0.7) +
    labs(x="Flight distance",
         y="Number of flights",
         color="Flight delayed or on-time?",
         title="Number of flights for each airline by distance and whether delayed or not")
```

... but flight distance itself is not a good predictor for delay, albeit it appears that longer-distance flights might be less likely delayed:
```{r}
#| fig.height: 10
#| fig.width: 16

dat |>
    ggplot() +
    geom_point(aes(x=distance, y=arr_delay, color=arr_delay_fct), alpha=0.2) +
    facet_wrap(~paste("Airline:", carrier), scales="free") +
    scale_color_viridis_d(option = "plasma", end=0.7) +
    labs(x="Flight distance",
         y="Arrival delay (minutes)",
         color="Flight delayed or on-time?",
         title="Arrival delay by airline and flight distance")
```

## Build a logistic regression model

```{r}
set.seed(123)
dat_split <- initial_split(select(dat, -arr_delay), prop=3/4)
dat_train <- training(dat_split)
dat_test <- testing(dat_split)
```

```{r}
dat_recipe <- recipe(arr_delay_fct ~ ., data=dat_train) |>
    update_role(flight, time_hour, new_role="identity") |>
    step_date(flight_date, features = c("dow", "month")) |>
    step_holiday(flight_date, 
                 holidays = timeDate::listHolidays("US"), 
                 keep_original_cols = FALSE) |>
    step_dummy(all_nominal_predictors()) |> # create dummy numerical variables out of factor columns
    step_zv(all_predictors()) # remove predictors with zero variation, i.e. single-valued columns

print(dat_recipe)
```

```{r}
model1 <- logistic_reg() |>
    set_engine("glm")

fitting_workflow1 <- workflow() |>
    add_model(model1) |>
    add_recipe(dat_recipe)

print(fitting_workflow1)
```

**Apply the model workflow on training data:**
```{r}
fitted_model1 <- fitting_workflow1 |>
    fit(dat_train)
```

```{r}
fitted_model1 |>
    extract_recipe()
```

```{r}
fitted_model1 |>
    extract_fit_engine() |>
    tidy()
```

**Run the fitted model on test data, and evaluate performance:**
```{r}
dat_test_with_predictions <- augment(fitted_model1, dat_test)
```

```{r}
glimpse(dat_test_with_predictions)

dat_test_with_predictions |> roc_auc(truth=arr_delay_fct, .pred_delayed)
```

```{r}
dat_test_with_predictions |>
    roc_curve(truth=arr_delay_fct, .pred_delayed) |>
    ggplot() +
    geom_path(aes(x=1-specificity, y=sensitivity), color="deeppink") +
    geom_abline(slope=1, intercept=0, color="black", linetype=2) +
    labs(x="False positive rate (propotion of flights incorrectly predicted as delayed)",
         y="True positive rate (propotion of flights correctly predicted as delayed)",
         title="ROC curve for model1")
```